% ============================================================================
% ABSTRACT — Journal of Big Data (expanded, max 350 words)
% ============================================================================
\begin{abstract}
Large Language Models (LLMs) have achieved strong performance on text-to-SQL
benchmarks, yet their ability to reason about the \emph{structural topology}
of database schemas remains unexplored.  Data warehouse environments demand
a fundamentally different class of reasoning: understanding foreign key paths,
data lineage chains, connected components, and multi-hop impact propagation
across heterogeneous schema graphs.  We introduce \textbf{DW-Bench}, the
first benchmark designed to evaluate LLMs on graph topology reasoning over
data warehouse schemas.

DW-Bench is organized in two tiers.  \textbf{Tier~1} (schema-level,
1,046~questions across 13~subtypes and 3~difficulty levels) tests structural
reasoning over foreign key paths, lineage chains, and connected component
detection using five datasets---four real-world (AdventureWorks, TPC-DS,
TPC-DI, OMOP~CDM) and one synthetic (Syn-Logistics) encompassing
262~tables and 521~edges.  \textbf{Tier~2} (value-level, 433~questions
across 8~subtypes) extends evaluation to row-level provenance tracing and
forward impact analysis, requiring interactive data access beyond any
context window.

We evaluate six baselines spanning the text-only to agentic spectrum:
Flat Text, Vector-RAG, Graph-Augmented, Tool-Use, ReAct-Code, and Oracle,
with two frontier LLMs (Gemini~2.5 Flash and DeepSeek-V3).  Our results
reveal four key insights: (1)~Tool-Use dominates schema-level reasoning
with 89--90\% micro-EM versus 69--77\% for static baselines; (2)~static
baselines exhibit a ``triviality illusion'' where easy subtypes inflate
aggregate scores; (3)~even tool-augmented LLMs plateau at $\sim$60\% on
hard multi-hop compositional reasoning (versus $>$97\% Oracle), identifying
a fundamental bottleneck; and (4)~on Tier~2 value-level questions, tools
are \emph{necessary} (100\% vs.\ 0\% on data-dependent subtypes) but
\emph{insufficient} for complex multi-hop reasoning (27\% on
\texttt{multi\_hop\_trace}), suggesting that learned structural
representations (e.g., GNN encoders) may be needed to close the gap.
An obfuscation protocol confirms that Tool-Use operates algorithmically
($\Delta{=}{-}3.4$\%) while static baselines lose 15--32\% EM.

All datasets, evaluation code, and baseline implementations are
publicly released to catalyze research on hybrid graph reasoning systems.
\end{abstract}
