% ============================================================================
% 6. DISCUSSION — Journal version (expanded)
% ============================================================================
\section{Discussion}
\label{sec:discussion}

\subsection{The Retrieval--Reasoning Decomposition}

Our results reveal a clean decomposition of graph reasoning performance
into two independent failure modes.

\textbf{Retrieval-limited subtypes} (\texttt{hop\_count},
\texttt{direct\_fk}, \texttt{isolation}, \texttt{connected}) achieve
$\geq$93\% EM with Tool-Use because the answer requires only one or
two graph lookups.  Static baselines fail on these subtypes
\emph{not because the reasoning is complex}, but because the relevant
graph fragment is missing from the injected context.  This is confirmed
by the Oracle's near-perfect scores: once the correct information is
available, both models can interpret it accurately.

\textbf{Reasoning-limited subtypes} (\texttt{combined\_impact},
\texttt{membership}, \texttt{transitive}) remain hard even with tools.
Even with Tool-Use providing graph algorithm access,
\texttt{combined\_impact} achieves only 12\% EM ($n=69$).  The LLM
struggles to compose multiple tool outputs across phases, confirming
a ceiling for prompt-based graph reasoning.  Notably, RC
(5~code execution rounds) shows a similar ceiling to TU (3~tool calls),
suggesting the bottleneck is compositional reasoning itself rather
than interaction budget.

\subsection{The Tool Necessity-Insufficiency Paradox}

Tier~2 reveals a striking paradox: tools are simultaneously
\emph{essential} and \emph{insufficient}.

On data-dependent subtypes (\texttt{cascade\_count}: 100\% vs.\ 0\%;
\texttt{row\_provenance}: 92\% vs.\ 0\%), tools are the difference
between perfect performance and complete failure.  These subtypes
require traversing the 18MB row-level lineage map---far exceeding any
context window.

Yet on reasoning-intensive subtypes (\texttt{multi\_hop\_trace}: 27\%;
\texttt{shared\_source}: 23\%), tools provide \emph{no} advantage over
static baselines.  These subtypes require multi-hop graph traversal
and set-theoretic reasoning (finding common ancestors of two mart
tables)---capabilities that tools provide but LLMs cannot effectively
orchestrate.

\subsection{Implications for Hybrid Architectures}

The complementarity analysis confirms that Tool-Use and Graph-Augmented
fail on largely the \emph{same} questions: their union improves overall
EM by only 6 percentage points, leaving 35\% of questions unsolved by
either approach.  This suggest that the bottleneck is not retrieval
\emph{or} tool access, but the LLM's ability to perform compositional
structural reasoning.

This points toward hybrid architectures combining:
\begin{enumerate}
  \item \textbf{Tool-augmented retrieval} for data-access subtypes
    where simple lookups suffice.
  \item \textbf{Learned structural representations} (e.g., GNN encoders)
    that can capture multi-hop topology patterns beyond what prompt-based
    reasoning can achieve.
  \item \textbf{Compositional reasoning modules} that decompose complex
    graph queries into sequences of simpler operations.
\end{enumerate}

\subsection{Obfuscation and Memorization}

The obfuscation results provide a clean diagnostic for memorization.
Static baselines lose 15--32\% EM under obfuscation, confirming partial
reliance on memorized schema name associations.  Tool-Use's near-invariance
($\Delta{=}{-}3.4$\% Gemini, ${-}4.2$\% DeepSeek) demonstrates that
algorithmic graph access is robust to input perturbation---a desirable
property for production deployment where schemas are diverse and
evolving.

\subsection{Model Differences}

Gemini and DeepSeek show consistent patterns but differ in magnitude.
DeepSeek-V3 benefits more from explicit graph context (GA outperforms
FT by 7.3pp for DeepSeek vs.\ only -1.1pp for Gemini), suggesting
that models with less domain exposure rely more heavily on structured
context injection.

\subsection{Positioning Against Stronger Agentic Methods}

Our baselines intentionally span the static-to-agentic spectrum rather
than exhaustively covering the agentic design space.  More sophisticated
approaches---planning-based agents with dynamic loop budgets (e.g.,
ToG-style multi-agent planners~\cite{sun2024tog}), hierarchical
GraphRAG~\cite{graphrag2024} with iterative subgraph expansion, or modular
query decomposition pipelines---may partially close the gap on compositional
subtypes.  However, the convergence of TU (3~calls) and RC (5~code rounds) at
similar hard-question ceilings suggests that the bottleneck is the
LLM's compositional reasoning capacity rather than the tool interface.
DW-Bench provides the diagnostic infrastructure for evaluating these
stronger methods as they emerge.

\subsection{The Oracle as Classical Solver Proxy}

A natural question is whether a non-LLM classical solver should be
included as a baseline.  Our Oracle baseline serves this role: it
injects gold algorithmic outputs (computed by NetworkX graph algorithms)
directly into the LLM prompt, so that the LLM need only transcribe and
format the answer.  The 3\% gap to 100\% EM is attributable to output
formatting (list ordering canonicalization, alternative valid shortest
paths)---not reasoning failures.  A fully classical pipeline would
achieve~100\% by bypassing the LLM entirely, confirming that the
bottleneck is LLM comprehension of compositional graph structure, not
algorithmic difficulty.

\paragraph{Formal specification of \texttt{combined\_impact}.}
This subtype computes the set of tables affected by a change to a source
table through \emph{both} lineage and FK paths:
\[
\textit{Impact}(t) = \textit{Down}_{\textit{lin}}(t)
\;\cup\; \bigcup_{t' \in \textit{Down}_{\textit{lin}}(t)}
\textit{Reach}_{\textit{FK}}(t'),
\]
where downstream lineage is transitive closure over \texttt{derived\_from}
edges and FK reachability is BFS over \texttt{fk\_to} edges.  The
composition of two distinct traversals is what makes this subtype hard:
TU achieves only 12\% EM despite having both \texttt{get\_lineage\_chain}
and \texttt{shortest\_path} tools available.

\subsection{GA Obfuscation Penalty Analysis}

The large GA obfuscation penalty ($\Delta{=}{-}27.9$\% Gemini,
${-}24.3$\% DeepSeek) may appear surprising given that BFS anchor tables
remain present after obfuscation.  We attribute this to two compounding
factors: (1)~the LLM loses semantic priors about table roles when
names like \texttt{DimCustomer} become \texttt{Table\_K}, making it
harder to interpret the 3-hop BFS neighborhood; and (2)~45\% of
questions require paths exceeding 3 hops, so GA already has incomplete
context---obfuscation compounds this by removing the semantic cues that
help LLMs \emph{infer} structure beyond the retrieved fragment.
In contrast, TU ($\Delta{=}{-}3.4$\%) and RC ($\Delta{=}{+}0.6$\%)
are robust because they access the graph algorithmically rather than
relying on name semantics.

\subsection{Tool Tautology and API Design}

The reviewer correctly observes that TU's high scores on easy subtypes
partly reflect ``API selection skill'' rather than deep reasoning---e.g.,
answering \texttt{hop\_count} by calling \texttt{shortest\_path} is
structurally straightforward.  We view this as a feature, not a limitation:
TU's near-perfect easy scores \emph{isolate} the hard-task failures as
genuine reasoning gaps.  If TU struggled on easy tasks too, it would be
unclear whether the hard-task ceiling reflects reasoning difficulty or
API design.  The fact that TU solves easy tasks trivially but fails on
\texttt{combined\_impact} (12\%) and \texttt{membership} (32\%)
demonstrates that the gap is compositional reasoning, not tool availability.

\subsection{Limitations}

We do not enforce token-budget parity across baselines by design:
each paradigm receives context natural to its architecture (full schema
for FT, retrieved chunks for VR, BFS subgraph for GA, tool results
for TU).  45\% of questions exceed the GA 3-hop radius.  We evaluate
two frontier models; broader model coverage (GPT-4, open-weight LLMs)
would strengthen generality.  Tier~2 is limited to one synthetic dataset;
extending to real datasets with observed lineage noise is planned.
All lineage edges are assumed to form DAGs; cyclic ETL patterns
(e.g., slowly-changing dimensions, feedback loops) are left for future
versions.  We deliberately exclude GNN-enhanced baselines: DW-Bench
evaluates LLM graph reasoning capabilities, not methods.  The
per-subtype EM diagnostics and obfuscation protocol provide the
infrastructure needed to evaluate GNN-augmented retrieval when such
approaches emerge.
Potential risks include overreliance on automated impact
analysis in production; careful validation remains essential.
