\begin{abstract}
Large Language Models (LLMs) have demonstrated remarkable capabilities on
text-to-SQL benchmarks such as Spider and BIRD. However, these benchmarks
evaluate \emph{query generation}, not \emph{structural understanding} of
database schemas. We introduce \textbf{DW-Bench}, the first benchmark
specifically designed to evaluate LLMs on \emph{graph topology reasoning}
over data warehouse schemas. DW-Bench contains \textbf{1,042 questions}
across \textbf{5 datasets} (4 real-world: AdventureWorks, TPC-DS, TPC-DI,
OMOP~CDM; 1 synthetic: Syn-Logistics) spanning \textbf{262 tables},
\textbf{13 question subtypes}, and \textbf{3 difficulty levels}. Questions
require reasoning about foreign key paths, data lineage chains, transitive
impact analysis, and connected component detection---tasks that demand
graph traversal rather than SQL generation.

We evaluate four baselines---\emph{Oracle}, \emph{Flat Text},
\emph{Vector-RAG}, and \emph{Graph-Augmented}---plus a \emph{Tool-Use}
baseline, using two frontier models: Gemini~2.5 Flash~\cite{gemini2024}
and DeepSeek-V3~\cite{deepseek2024v3}. Gemini achieves strong performance
on structural subtypes (\texttt{hop\_count}: 98--100\% EM,
\texttt{count}: 100\% EM), demonstrating that frontier models
\emph{can} perform schema topology reasoning when the schema fits in
context. However, the \textbf{model-dependent gap} is striking:
DeepSeek-V3 scores only 48\% on \texttt{hop\_count} (FT) and 27\% on
\texttt{count} (VR), revealing that structural reasoning capability
varies dramatically across architectures. On \texttt{combined\_impact}
($n{=}34$), \emph{all} baselines score $\leq$26.5\% while the Oracle
achieves 100\%, identifying a genuine frontier for multi-hop structural
reasoning.

Micro-averaging creates a \textbf{Triviality Illusion}: easy subtypes
inflate scores, masking failures on hard structural tasks that
macro-averaging reveals. DW-Bench includes an obfuscation protocol
replacing table names with random identifiers, showing DeepSeek suffers
up to \textbf{30\%} penalty while Gemini shows differential effects.
The benchmark, all results, and evaluation code are publicly available at
\url{https://github.com/AJamal27891/dw-bench}.
\end{abstract}
