% ============================================================================
% APPENDIX
% ============================================================================
\appendix
\section{Dataset Details}
\label{app:datasets}

% FIGURE A1: Full schema graphs for all 4 datasets
% [FIGURE A1 PLACEHOLDER: 2x2 grid of schema graph visualizations]

\paragraph{Syn-Logistics Design.}
The Syn-Logistics dataset was constructed to guarantee $n\geq20$ for every
subtype by building three parallel 4-hop lineage chains across five
connected components. Table names are drawn from a Domain Dictionary of
realistic supply-chain terminology to prevent semantic collapse under
obfuscation. The schema contains 64 tables, 96 FK edges, 35 lineage edges,
and 372 questions.

\paragraph{Lineage Edge Curation.}
Lineage (derived\_from) edges are curated as follows:
\textbf{AdventureWorks}: Extracted from official Microsoft reference
documentation mapping OLTP source tables to DW dimension/fact tables (39
edges).
\textbf{TPC-DS}: No lineage edges; schema is a pure star/snowflake FK
graph.
\textbf{TPC-DI}: Derived from the official TPC-DI specification defining
staging$\to$dimension$\to$fact transformation flows (21 edges).
\textbf{OMOP CDM}: Derived from OHDSI ETL convention guides mapping
source vocabularies and clinical tables to standardized CDM tables (21
edges).
\textbf{Syn-Logistics}: Defined by construction---three parallel 4-hop
chains (raw$\to$staging$\to$dimension$\to$mart) per connected component
(35 edges). All lineage edges are validated by verifying that the source
table's schema semantically feeds the target table's schema.

\section{Hop-Distance Analysis}
\label{app:hop-distance}

\begin{table}[h]
\centering
\caption{Minimum evidence hops per subtype on Syn-Logistics, compared to
Graph-Augmented (Gemini) EM. Subtypes requiring full-graph knowledge
(silo detection, component enumeration) inherently exceed the 3-hop BFS
radius. Overall, \textbf{45.4\%} of questions exceed 3 hops.}
\label{tab:hop-distance}
\small
\begin{tabular}{lrrrc}
\toprule
Subtype & $n$ & Exceeds 3h & \% & GA EM \\
\midrule
count            & 26 & 26 & 100\% & 100\% \\
membership       & 20 & 20 & 100\% & 0\% \\
combined\_impact & 34 & 34 & 100\% & 0\% \\
full\_enumeration & 25 & 25 & 100\% & 100\% \\
connected        & 25 & 25 & 100\% & 100\% \\
isolation        & 25 & 25 & 100\% & 100\% \\
transitive       & 29 &  4 &  14\% & 93\% \\
hop\_count       & 44 &  5 &  11\% & 100\% \\
direct\_fk       & 36 &  3 &   8\% & 100\% \\
join\_path       & 26 &  2 &   8\% & 81\% \\
forward          & 20 &  0 &   0\% & 100\% \\
reverse          & 20 &  0 &   0\% & 100\% \\
multi\_source    & 42 &  0 &   0\% & 95\% \\
\midrule
\textbf{Total}   & 372 & 169 & \textbf{45.4\%} & 83.1\% \\
\bottomrule
\end{tabular}
\end{table}

\section{Full Per-Subtype Results}
\label{app:subtypes}

% TABLE A1: Full Syn-Logistics per-subtype table (CORRECTED)
\begin{table}[h]
\centering
\caption{Per-subtype EM (\%) on Syn-Logistics (corrected). Gemini achieves
near-perfect scores on \texttt{hop\_count} and \texttt{count}, while
DeepSeek shows a pronounced model-dependent gap. Both models
struggle on \texttt{combined\_impact} and \texttt{membership}.}
\label{tab:syn-subtypes}
\small
\begin{tabular}{lrcccccc}
\toprule
Subtype & $n$ & \multicolumn{3}{c}{Gemini 2.5 Flash} & \multicolumn{3}{c}{DeepSeek-V3} \\
\cmidrule(lr){3-5} \cmidrule(lr){6-8}
 & & FT & GA & VR & FT & GA & VR \\
\midrule
hop\_count       & 44 & 98 & 100 & 93 & 48 & 100 & 70 \\
count            & 26 & 100 & 100 & 100 & 69 & 73 & 27 \\
combined\_impact & 34 & 2.9 & 0 & 0 & 26.5 & 2.9 & 17.6 \\
membership       & 20 & 25 & 0 & 25 & 40 & 0 & 20 \\
\midrule
forward          & 20 & 100 & 100 & 100 & 60 & 35 & 50 \\
reverse          & 20 & 100 & 100 & 90 & 75 & 55 & 60 \\
transitive       & 29 & 96.6 & 93.1 & 100 & 72.4 & 69 & 75.9 \\
multi\_source    & 42 & 100 & 95.2 & 81 & 92.9 & 57.1 & 83.3 \\
join\_path       & 26 & 100 & 80.8 & 92.3 & 88.5 & 61.5 & 84.6 \\
full\_enum       & 25 & 100 & 100 & 72 & 88 & 80 & 56 \\
connected        & 25 & 100 & 100 & 100 & 100 & 100 & 84 \\
isolation        & 25 & 100 & 100 & 100 & 100 & 100 & 100 \\
direct\_fk       & 36 & 100 & 100 & 100 & 100 & 100 & 100 \\
\bottomrule
\end{tabular}
\end{table}

\section{Decoding and Evaluation Details}
\label{app:decoding}

All LLM inferences use greedy decoding ($\tau{=}0.0$) with no nucleus
sampling ($p = 1.0$) and a single deterministic pass. No repeated runs or
confidence intervals are reported; the deterministic decoding ensures
reproducibility. For Vector-RAG, we use
\texttt{all-MiniLM-L6-v2}~\cite{reimers2019sentencebert} with $k{=}15$
retrieved chunks per question. List answers are compared as unordered sets
after lowercasing; exact match requires identical sorted sets. F1 is
computed as token-level overlap between predicted and gold answer sets.

\section{Tool-Use Baseline}
\label{app:tool-use}

The Tool-Use baseline provides the LLM with 9 graph algorithm tools
(\texttt{shortest\_path}, \texttt{connected\_components},
\texttt{check\_fk\_adjacency}, \texttt{get\_lineage\_forward},
\texttt{get\_lineage\_reverse}, \texttt{transitive\_lineage},
\texttt{get\_component\_of}, \texttt{get\_fk\_neighbors},
\texttt{list\_tables}) implemented via NetworkX. The LLM receives a
schema summary (table list + edge types) and tool definitions, then
generates up to 3 tool calls per question in a multi-turn conversation.
Tool results are returned as JSON and the LLM formulates its final answer.
Gemini Tool-Use achieves 86.8\% Micro-EM on Syn-Logistics, matching Flat
Text (86.6\%), confirming that the LLM can reason directly from textual
schema descriptions without needing explicit tool access for small schemas.

\section{Example Questions}
\label{app:examples}

\paragraph{Easy (direct\_fk):}
\emph{``Is there a direct foreign key relationship between
\texttt{DW.DimProduct} and \texttt{DW.FactInternetSales}?''}
\textbf{Gold: yes}

\paragraph{Medium (join\_path):}
\emph{``What is the shortest FK path from \texttt{DW.DimCustomer} to
\texttt{DW.DimProduct}?''}
\textbf{Gold: [DW.DimCustomer, DW.FactInternetSales, DW.DimProduct]}

\paragraph{Hard (combined\_impact):}
\emph{``Which tables are transitively affected (via lineage and FK
dependencies) if \texttt{raw\_purchase\_orders} is modified?''}
\textbf{Gold: [stg\_purchase\_orders, dim\_supplier, ...]} (8 tables)
