% ============================================================================
% 6. CONCLUSION
% ============================================================================
\section{Conclusion}

DW-Bench tests whether LLMs can reason about the graph topology of data
warehouse schemas.  We evaluate six baselines across five datasets using
three models, yielding 1,046 schema-level and 953 value-level questions.
Four findings stand out.

\textbf{(1)~Tool-use dominance.}  Give the LLM deterministic graph tools
and it reaches 87--92\% micro-EM, beating the best static baseline by
7--12~pp.  LLMs can orchestrate algorithms they cannot internally simulate.

\textbf{(2)~Triviality illusion.}  Easy lookup subtypes inflate micro-EM.
Switch to macro-EM (equal subtype weight) and the ranking shifts: what
looked like widespread success reveals significant performance disparities.

\textbf{(3)~Hard-task ceiling.}  No baseline, including agentic
code-generation, exceeds 61\% on hard questions; the Oracle hits 100\%.
One subtype, \texttt{combined\_impact} (12\% EM with tools), accounts for
most of the gap: it requires chaining lineage traversal with FK expansion,
a compositional step that current baselines handle poorly.

\textbf{(4)~Obfuscation invariance.}  Tool-Use barely moves under
obfuscation ($\Delta < 4$~pp).  Static baselines drop 9--32~pp, consistent
with tools operating on topology rather than memorized table names.

The natural next step is a \textbf{hybrid architecture}: pair interactive
tools with learned graph representations (GNN encoders, graph-aware
controllers) to close the compositional reasoning gap.  DW-Bench supplies
the per-subtype diagnostics needed to measure progress.
