% ============================================================================
% 6. DISCUSSION
% ============================================================================
\section{Discussion}

\paragraph{The model-dependent structural gap.}
The most striking finding is the dramatic performance difference between
Gemini and DeepSeek on structural subtypes. Gemini achieves 98--100\% EM
on \texttt{hop\_count} and 100\% on \texttt{count} across FT, GA, and VR,
while DeepSeek scores as low as 48\% and 27\% respectively. This
suggests that structural graph reasoning capability is not uniformly
present in frontier models but depends on architecture and training
data composition. DW-Bench enables precise quantification of this gap.

\paragraph{Why does Flat Text win on small schemas?}
On Syn-Logistics (64 tables), Flat Text (86.6\%) outperforms Graph-Augmented
(83.1\%) because the complete schema fits in the LLM's context
window---Gemini's 1M-token context easily accommodates a 64-table schema.
For local queries (direct FK, forward lineage), the LLM can
pattern-match the answer from text. On larger schemas (102+ tables),
GA's targeted retrieval becomes essential as context dilution degrades FT.

\paragraph{Why does Vector RAG underperform?}
Chunking the schema into per-table documents fragments the graph,
breaking cross-chunk edges. Despite our global lineage summary
mitigation, VR still misses schema-wide connectivity information,
performing worst on \texttt{combined\_impact} and \texttt{membership}.

\paragraph{The combined\_impact frontier.}
This subtype requires two distinct traversals: (1) forward lineage from
a source table, collecting all transitively-affected DW tables, then (2)
reverse FK from each affected table, finding all tables that depend on
them. With $\leq$26.5\% EM across both models and all baselines, while
Oracle achieves 100\%, \texttt{combined\_impact} represents the genuine
unsolved frontier---even when the schema fits in context and tool
access is provided.

\paragraph{Limitations.}
DW-Bench currently evaluates schema-level topology reasoning on schemas
small enough to fit in context (24--102 tables). Real enterprise data
warehouses can have thousands of tables with millions of data rows;
extending DW-Bench to test data-level graph reasoning (with value nodes)
at scale is an important direction. The benchmark includes 5 datasets
(4 real-world, 1 synthetic); more diverse topologies would strengthen
generalization claims.
