% ============================================================================
% 5. EXPERIMENTS
% ============================================================================
\section{Experiments}

\subsection{Setup}

We evaluate using Gemini 2.5 Flash via API. Each question receives the
baseline-specific context and returns a JSON response with reasoning and
answer. We measure:

\begin{itemize}
  \item \textbf{Exact Match (EM)}: Binary correctness. For list answers,
    we compare sorted sets.
  \item \textbf{F1 Score}: Token-level overlap for partial credit on list
    answers.
\end{itemize}

\subsection{Main Results}

% TABLE 3: Main Results
\begin{table}[t]
\centering
\caption{Main results: Exact Match (\%) across datasets and baselines
using Gemini 2.5 Flash. \textbf{Bold} = best per column.}
\label{tab:main-results}
\begin{tabular}{lccccc}
\toprule
Baseline & AW & TPC-DS & TPC-DI & OMOP & \textbf{Avg} \\
\midrule
Flat Text   & 76.0 & \textbf{81.9} & \textbf{76.2} & \textbf{62.0} & \textbf{74.0} \\
Vector RAG  & 68.1 & 69.3 & 74.6 & 60.8 & 68.2 \\
Graph-Aug   & \textbf{77.5} & 70.1 & 71.8 & 60.1 & 69.9 \\
\bottomrule
\end{tabular}
\end{table}

% FIGURE 3: Bar chart grouped by difficulty
% [FIGURE 3 PLACEHOLDER: Three-group bar chart (Easy/Medium/Hard)
% with FT, VR, GA bars in each group. The Hard group should show
% GA clearly ahead.]

\subsection{Results by Difficulty}

% TABLE 4: By difficulty
\begin{table}[t]
\centering
\caption{Exact Match (\%) by difficulty level. All baselines
\textbf{collapse on hard} questions ($\leq$17.5\% EM).}
\label{tab:difficulty}
\begin{tabular}{lcccc}
\toprule
Difficulty & $n$ & FT & VR & GA \\
\midrule
Easy    & 226 & \textbf{83.2} & 81.9 & 78.4 \\
Medium  & 383 & \textbf{77.8} & 71.3 & 74.2 \\
Hard    &  61 & 14.8 & 5.3 & \textbf{17.5} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Key Finding 1: All baselines collapse on hard tasks.}
While Flat Text achieves the highest overall EM (74.0\%), all baselines
collapse on hard structural reasoning: FT drops to 14.8\%, VR to 5.3\%,
and GA to 17.5\%. Graph-Aug provides a modest +2.8\% improvement over FT
on hard tasks, but the absolute performance remains critically low.

\subsection{Per-Subtype Analysis}

% FIGURE 4: Heatmap of EM by subtype x baseline
% [FIGURE 4 PLACEHOLDER: 13-row x 3-column heatmap
% Rows = subtypes (sorted by difficulty), Cols = FT/VR/GA
% Color: green (100%) to red (0%)]

\paragraph{Key Finding 2: Specific subtypes reveal graph reasoning gaps.}
Both FT and GA achieve 100\% on \texttt{count}, \texttt{full\_enumeration},
and \texttt{multi\_source}---structurally simple tasks. GA achieves
50\% on \texttt{transitive} (multi-hop lineage) where FT scores 0\%,
and 50\% on \texttt{connected} where FT scores 25\%. FT leads on
\texttt{join\_path} (77.8\% vs.\ GA's 69.3\%) due to its complete FK
context.

\paragraph{Key Finding 3: Combined impact remains unsolved.}
The \texttt{combined\_impact} subtype---requiring simultaneous traversal
of lineage chains AND FK dependency trees---achieves $\leq$6.5\% EM across
all baselines (FT: 6.5\%, GA: 3.2\%). This establishes a clear frontier
for future work on learned graph representations.

\subsection{Obfuscation Results}

% TABLE 5: Original vs Obfuscated
\begin{table}[t]
\centering
\caption{Original vs.\ obfuscated exact match (\%). Large deltas indicate
reliance on schema name memorization rather than structural reasoning.}
\label{tab:obfuscation}
\begin{tabular}{lccc}
\toprule
Baseline & Original & Obfuscated & $\Delta$ \\
\midrule
Flat Text   & 73.9 & 64.9 & $-$9.0 \\
Vector RAG  & 68.4 & 50.1 & $-$18.2 \\
Graph-Aug   & 70.4 & 49.1 & $-$21.3 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Key Finding 4: Obfuscation reveals a dissociation.}
Easy question performance drops 25--33\% when table names are scrambled,
confirming that models rely on name recognition for routine structural
tasks (forward lineage, FK adjacency). In contrast, hard structural
subtypes score near 0\% \emph{regardless} of obfuscation, confirming
these failures reflect structural reasoning impossibilities rather than
knowledge gaps. The model cannot perform multi-hop traversal even when
it knows the real table names.

Graph-Aug suffers the largest obfuscation penalty ($-$21.3\%), suggesting
that its graph algorithm output is easier to interpret when table names
carry semantic meaning. Vector RAG's large penalty ($-$18.2\%) reflects
that embedding-based retrieval degrades when table name semantics are
removed.
