% ============================================================================
% 2. RELATED WORK
% ============================================================================
\section{Related Work}

\paragraph{Text-to-SQL Benchmarks.}
Spider~\cite{yu2018spider} introduced cross-database text-to-SQL evaluation
with 10,181 questions across 200 databases. BIRD~\cite{li2023bird} extended
this with noisy real-world schemas. Spider~2.0~\cite{lei2024spider2} further
increased difficulty with enterprise-scale schemas. ScienceBenchmark
evaluates SQL generation on scientific databases.
ToRR~\cite{torr2024} evaluates table reasoning robustness on cell/row
logic---complementary to DW-Bench's topology focus. All target
\emph{query generation or cell reasoning}, not structural understanding.

\paragraph{Schema Understanding.}
BEAVER~\cite{beaver2024} evaluates LLMs on database knowledge including
schema matching and entity resolution, but does not evaluate graph topology
reasoning.  LLM-FK~\cite{llmfk2024} evaluates FK discovery---a single-hop
structural task complementary to DW-Bench's multi-hop topology reasoning.

\paragraph{Graph Reasoning in LLMs.}
Recent work has explored LLM capabilities on graph tasks.
NLGraph~\cite{wang2023nlgraph} evaluates LLMs on basic graph problems
(connectivity, shortest path). GraphQA~\cite{fatemi2023talk} further
explores graph reasoning. However, these use synthetic random graphs, not
real-world database schemas with heterogeneous edge types (FK vs.\ lineage).
GraCoRe~\cite{gracore2024} provides a comprehensive taxonomy of graph
reasoning tasks, and GraphArena~\cite{grapharena2024} benchmarks graph
algorithms at scale---but neither addresses the domain-specific challenges
of data warehouse schemas with mixed FK and lineage edge types.

\paragraph{Advanced GraphRAG and Agentic Methods.}
Recent literature highlights advanced modular GraphRAG, RL-finetuning
(e.g., G1), and agentic tool-use (e.g., ToG-3.0, FABLE, GraphCogent,
GDS Agent) as promising directions for general graph traversal. However,
in alignment with foundational dataset methodologies (e.g., Spider, BIRD),
DW-Bench deliberately evaluates standardized, zero-shot baselines (Flat
Text, standard Vector-RAG) to establish the empirical lower-bound of LLM
structural reasoning. Hierarchical GraphRAG~\cite{graphrag2024} and
emerging RL-tuned retrieval orchestrators represent natural next steps.
By formalizing the evaluation target and providing
the exact Oracle upper-bound (98\%+ EM), we isolate the topological
retrieval bottleneck; closing this gap with composite, multi-agent
frameworks is explicitly left as future work.

\paragraph{Positioning.}
DW-Bench uniquely combines: (1) \emph{real-world} schemas from industry
standards, (2) \emph{heterogeneous} edge types (FK + lineage), (3)
\emph{practical} question types from data engineering workflows, and (4)
\emph{obfuscation} for contamination control.
